refnx - Neutron and X-ray reflectometry analysis in Python

### 2018-07-17 paper/manuscript.tex
PAPER: reproducibility comments

To make it easy to create quick models, and to ease the learning curve, an ipywidgets \cite{ipywidgets} based fitting GUI can be created in a \Jupyter notebook, Figure~\ref{fig:gui}. This GUI builds slab based models for analysing a single dataset, but can be used as a stepping point for building more advanced models via the 'To code' code generation button. A fully functional web-based reflectometry analysis notebook is currently available \cite{Nelson2018}. For a comprehensive demonstration of multiple contrast co-refinement see \textbf{lipid.ipynb}, \textbf{lipid.pdf} in the supporting information.

@@

\section*{Comments on reproducibility of analyses}
In order for a given scattering analysis to be fully reproducible by others a general set of conditions need to be met:
\begin{itemize}
  \item the datasets used need to be deposited with a journal article, or be freely available.
  \item the exact software environment needs to be recreateable.
  \item the exact ordered set of steps taken during the analysis needs to be listed.
\end{itemize} 
Each of these points are frequently neglected. For example, the use of different software versions may change the output of an analysis, or the use of a GUI program may not record the full set of steps, or options, applied by a user.
Whilst it is unable to meet the first criteria by itself, the use of \refnx in a \Jupyter notebook can fulfil the other two requirements, providing a little care is taken. The exact software environment can be recreated by noting down the versions of the software packages used during an analysis (\refnx, \SciPy, \NumPy, Python, etc). At a later date those exact versions can be installed in the same Python version using one of: the \conda package manager, by installing from the source at a given version tag in the git repository, or by \pip. \conda can use environment files to recreate a specific setup. An alternative way of recreating the environment is a virtual machine, such as Docker. The ordered set of steps to perform the analysis is recordable in the \Jupyter notebook.
The usefulness of open-source software in a git (or other version controlled) repository must be emphasised here - the ability to return to a specific software version/environment is often difficult with proprietary or closed source software; different versions can have modifications that can significantly change the output of an analysis. Moreover, there are important ramifications for verifiability. \refnx is based on a fully open software stack, with good unit test coverage. The user can run tests for each component and inspect parts for correctness. For example, the behaviour of the reflectivity calculation in \refnx is checked from first principles in the test suite; and can be done now and in several years time. If problems are discovered, they can be corrected. With a closed source program such checking is harder, one has to examine the inputs/outputs of a black box without knowing what is happening in between, making it harder to find errors; one has to rely on trust. Whilst this may not be an issue for relatively simple calculations, it may be for complex algorithms, such as those involved with minimisation.

>>>>

To make it easy to create quick models, and to ease the learning curve, an ipywidgets \cite{ipywidgets} based fitting graphical user interface (GUI) can be created in a \Jupyter notebook, Figure~\ref{fig:gui}. This GUI builds slab based models for analysing a single dataset, but can be used as a stepping point for building more advanced models via the 'To code' code generation button. A fully functional web-based reflectometry analysis notebook is currently available \cite{Nelson2018}. For a comprehensive demonstration of multiple contrast co-refinement see \textbf{lipid.ipynb}, \textbf{lipid.pdf} in the supporting information.

@@

\section*{Comments on reproducibility of analyses}
In order for a given scattering analysis to be fully reproducible by others a general set of conditions need to be met:
\begin{itemize}
  \item the datasets used need to be deposited with a journal article, or be freely available.
  \item the exact software environment needs to be recreatable.
  \item the exact ordered set of steps taken during the analysis needs to be listed.
\end{itemize} 
Each of these points are often omitted in the literature. For example, the use of different software versions may change the output of an analysis, or the use of a GUI program may not record the full set of steps, or options, applied by a user.
Whilst it is unable to meet the first criteria by itself, the use of \refnx in a \Jupyter notebook can fulfil the other two requirements, providing a little care is taken. The exact software environment can be recreated by noting down the versions of the software packages used during an analysis (\refnx, \SciPy, \NumPy, Python, etc). At a later date those exact versions can be installed in the same Python version using one of: the \conda package manager, by installing from the source at a given version tag in the git repository, or by \pip. \conda can use environment files to recreate a specific setup. An alternative way of recreating the environment is by using a virtual machine, such as Docker. The ordered set of steps to perform the analysis is recordable in the \Jupyter notebook.
The usefulness of open-source software in a git (or other version controlled) repository must be emphasised here. With closed source or proprietary software the ability to return to a specific software version/environment can be difficult, different versions can have modifications that can change the output of an analysis. In addition accessibility (due to cost, etc) to the wider scientific community can also hinder reproducibility.
Moreover, there are important ramifications for verifiability. \refnx is based on a fully open software stack, with good unit test coverage. The user can run tests for each component and inspect parts for correctness. For example, the behaviour of the reflectivity calculation in \refnx is checked from first principles in the test suite; and can be done now and in several years time. If problems are discovered, they can be corrected. With a closed source program such checking is harder, one has to examine the inputs/outputs of a black box without knowing what is happening in between, making it harder to find errors; one has to rely on trust. Whilst this may not be an issue for relatively simple calculations, it may be for complex algorithms, such as those involved with minimisation.
